Awareness and Russell's Epistemology

### 2011-01-30 foundationalism.tex
changed discussion of hintikka's refutation

Russell's theory of knowledge is presented in Chapter XIII, entitled
\emph{Knowledge, Error, and Probable Opinion} \cite[pg. 52]{russell_problems_1936}:
\begin{quote}
We may now take a survey of the sources of our knowledge, as they have
appeared in the course of our analysis. We have first to distinguish knowledge
of things and knowledge of truths. In each there are two kinds, one immediate
and one derivative. Our immediate knowledge of things, which we called
\emph{acquaintance}, consists of two sorts, according as the things known are
particulars or universals. Among particulars, we have acquaintance with
sense-data and (probably) with ourselves.%  Among universals, there seems to
% be no principle by which we can decide which can be known by acquaintance,
% but it is clear that among those that can be so known are sensible qualities,
% relations of space and time, similarity, and certain abstract
% logical universals.
\ldots Our derivative knowledge of things, which we call knowledge by \emph{description},
always involves both acquaintance with something and knowledge of truths.
Our immediate knowledge of truths may be called \emph{intuitive} knowledge, and the
truths so known may be called \emph{self-evident} truths. Among such truths are
included those which merely state what is given in sense, and also certain
abstract logical and arithmetical principles, and (though with less certainty)
some ethical propositions. Our derivative knowledge of truths consists of
everything that we can deduce from self-evident truths by the use of self-
evident principles of deduction.
\end{quote}

@@

Historically, Hintikka briefly entertains a similar reading of
epistemic logic in \emph{Knowledge and Belief}
\cite{hintikka_knowledge_1969}[pgs. 38--39].  He proposes that ``$K_a
p$" might be read as ``It follows from what $a$ knows that $p$.''
Hintikka then rejects this reading on the grounds that it appeals to
some given standard of logical consequence.  He asserts that this is
``diametrically opposed'' to his research program in epistemic logic,
which is intended to ascribe a definition of logical consequence in
terms of the semantics he invents.

Hintikka is too dismissive of the foundationalist reading
of epistemic logic.  While unintuitive from a semantic
standpoint, it is possible to enforce a reading of knowledge in terms
of logical consequence. In the next section we sketch how this may be
accomplished, by employing awareness
logic and following Russell's formulation of knowledge by description.

@@

As we discussed in \S\ref{Russell}, \emph{Acquaintance with $\phi$} is
naturally interpreted as denoted as \emph{explicit awareness of
  $\phi$}.  We will denote this as $A:\phi$, with the usual
semantics of $\mathbb{M},w \models A:\phi \iff \phi \in \mathcal{A}_w$
where $\mathcal{A}_w$ is an awareness set.

@@

All that is left is to establish a connection between knowledge by
description and logical
consequence.  Let $\vdash$ be any Hilbert-style calculus,
that is a sequent that takes a set of premises $\Gamma$ on the left and yields a
single conclusion $\phi$ on the right. Define the following terminology:
\begin{itemizedot}
\item $\vdash$ is \emph{relatively sound} for $\mathbb{M}$ if and only if
  when $\Gamma \vdash \phi$, then if $\mathbb{M},w\models \gamma$, for
  all $\gamma \in \Gamma$, then
  $\mathbb{M},w \models \phi$.
\item $\vdash$ is \emph{reflexive} if and only if whenever $\phi \in
  \Gamma$ then $\Gamma \vdash \phi$
\item $\vdash$ makes true \emph{modus ponens} if and only if whenever
  $\Gamma \vdash \phi \to \psi$ and $\Gamma \vdash \phi$, then $\Gamma
  \vdash \psi$
\item $\vdash$ is said to be \emph{suitable} for $\mathbb{M}$ just in
  case that it is reflexive, makes true modus ponens, and is
  relatively sound for $\mathbb{M}$.
\end{itemizedot}

@@

\[ \mathbb{M}, w \models \Box \phi \textup{ if and only if } Th(\mathbb{M}) \cup
\mathcal{A}_w \vdash \phi \]

>>>>

Russell's theory of knowledge is a form of \emph{strong
  foundationalism}.  We specifically focus on the view presented in
Chapter XIII of \emph{The Problems of Philosophy}, entitled
``Knowledge, Error, and Probable Opinion'' \cite[pg. 52]{russell_problems_1936}:
\begin{quote}
We may now take a survey of the sources of our knowledge, as they have
appeared in the course of our analysis. We have first to distinguish knowledge
of things and knowledge of truths. In each there are two kinds, one immediate
and one derivative. Our immediate knowledge of things, which we called
\emph{acquaintance}, consists of two sorts, according as the things known are
particulars or universals. Among particulars, we have acquaintance with
sense-data and (probably) with ourselves.%  Among universals, there seems to
% be no principle by which we can decide which can be known by acquaintance,
% but it is clear that among those that can be so known are sensible qualities,
% relations of space and time, similarity, and certain abstract
% logical universals.
\ldots Our derivative knowledge of things, which we call knowledge by \emph{description},
always involves both acquaintance with something and knowledge of truths.
Our immediate knowledge of truths may be called \emph{intuitive} knowledge, and the
truths so known may be called \emph{self-evident} truths. Among such truths are
included those which merely state what is given in sense, and also certain
abstract logical and arithmetical principles, and (though with less certainty)
some ethical propositions. Our derivative knowledge of truths consists of
everything that we can deduce from self-evident truths by the use of self-
evident principles of deduction.
\end{quote}

@@

\subsection{Jaako Hintika}\label{Russell}

Jaako Hintikka briefly entertains a reading of
epistemic logic similar to Russell's epistemology 
in \emph{Knowledge and Belief}, where he provides the
first formulation of epistemic logic
\cite{hintikka_knowledge_1969}[pgs. 38--39].  
There he proposes that ``$K_a
p$" might be read as ``It follows from what $a$ knows that $p$''.

Hintikka rejects the Russellian reading of epistemic logic after
proposing it. His stated grounds are that
such a reading appeals to some given standard of logical consequence.  
He asserts that this is ``diametrically opposed'' to his research program in epistemic logic,
which is intended to develop a system of logical consequence using his
invented semantics as a foundation.

Hintikka's objection is reasonable.  It is counter intuitive to assume
a system of logical consequence is present at the semantic level in
epistemic logic.
However, it is not \emph{impossible}.  In the next section we sketch
an approach to epistemic logic that incorporates logical consequence
into awareness logic.  The intuition behind this move follows
Russell's formulation of \emph{acquaintance} and \emph{knowledge by
  description}.  After providing formal systems that express our
proposed reading of epistemic modality, we return Hintikka's objection
in \S\ref{hintikka-obj} in our concluding remarks.

@@

As we discussed in \S\ref{Russell}, \emph{Acquaintance with $\phi$} is
naturally interpreted as denoted as \emph{explicit awareness of
  $\phi$}.  We will denote this as $A:\phi$, with the usual
semantics of $\mathbb{M},w \models A:\phi \iff \phi \in \mathcal{A}(w)$
where $\mathcal{A}(w)$ is an awareness set.

@@

All that is left is to establish a connection between knowledge by
description and logical
consequence.  Let $\vdash$ be any Hilbert-style calculus,
that is a sequent that takes a set of premises $\Gamma$ on the left and yields a
single conclusion $\phi$ on the right. Define the following terminology:
\begin{definition}\ 
\begin{itemizedot}
\item $\vdash$ is \emph{relatively sound} for $\mathbb{M}$ if and only if
  when $\Gamma \vdash \phi$, then if $\mathbb{M},w\models \gamma$, for
  all $\gamma \in \Gamma$, then
  $\mathbb{M},w \models \phi$
\item $\vdash$ is \emph{reflexive} if and only if whenever $\phi \in
  \Gamma$ then $\Gamma \vdash \phi$
\item $\vdash$ makes true \emph{modus ponens} if and only if whenever
  $\Gamma \vdash \phi \to \psi$ and $\Gamma \vdash \phi$, then $\Gamma
  \vdash \psi$
\item $\vdash$ is said to be \emph{suitable} for $\mathbb{M}$ just in
  case that it is reflexive, makes true modus ponens, and is
  relatively sound for $\mathbb{M}$
\end{itemizedot}
\end{definition}

@@

\[ \mathbb{M}, w \models \Box \phi \textup{ if and only if } Th(\mathbb{M}) \cup
\mathcal{A}(w) \vdash \phi \]
